---
title: "Tidying_data"
author: "ayandabuso"
date: "2025-02-20"
output: html_document
---
#Introduction 
In this document we will tidy the insect data which is a mock data for my honours project.
#loading data
This data is already in my working directiory which i just saved directly to after creating it.Fistly, we load necessary libraries to read in data.

```{r}
#loading necessary libraries
library(tidyverse)
#we need to also add package to read an excel file
# Read the CSV file
insect_data <- read_csv("insects_data.csv")
# Check the first few rows
head(insect_data)

```
Now we are going to check our data and look for missing values

```{r}

# check for missing values
sum(is.na(insect_data))
```
I would like my data to to use proper colomn names for easier understanding of what they mean.

```{r}
# Convert data types
insect_data <- insect_data %>%
  mutate(
    pitfall_id = as.factor(pitfall_id),
    site_name = as.factor(site),
    collection_date = as.Date(date),
    insect_taxa = as.factor(taxa),
    count = as.numeric(count)  # Ensure count is numeric
  )

# Check structure
glimpse(insect_data)
```
We are interested in knowing the total abundance of each insect group recorded in the"taxa" colomn which is iportant in addressing the question, the code below answeres that.

```{r}
taxa_abundance <- insect_data %>%
  group_by(insect_taxa) %>%
  summarise(total_abundance = sum(count, na.rm = TRUE)) %>%
  arrange(desc(total_abundance))

# View summarized table
print(taxa_abundance)
```
lets visualise these abundance data!
```{r}
#first we need to load ggplot2 for this visualization.
library(ggplot2)
# Create a bar plot of insect abundance
ggplot(taxa_abundance, aes(x = reorder(insect_taxa, -total_abundance), y = total_abundance, fill = insect_taxa)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Abundance of Each Insect Taxa",
       x = "Insect Taxa",
       y = "Total Count") +
  theme_minimal() +
  theme(legend.position = "none") +
  coord_flip()
```
beatiful!
Now that we have gotten some knowledge about the abundance of the insect groups for example noting that the ants are most abundant and the flies are the least, but this barplot also helps us easily answer the second question of interest which is the diversity! we can tell that there are six different groups of insects found; flies,grasshoppers, bettles, thrips,butterflies and ants.

we can compare diversity between sites:

```{r}
# Calculate diversity by counting unique taxa per site
site_diversity <- insect_data %>%
  group_by(site_name) %>%
  summarise(diversity = n_distinct(insect_taxa)) %>%
  arrange(desc(diversity))

# View results
print(site_diversity)
```

The overall diversity:
```{r}
# Calculate the total number of unique insect taxa across all sites
overall_diversity <- insect_data %>%
  summarise(total_unique_taxa = n_distinct(insect_taxa))

# Print result
print(overall_diversity)
```

```{r}
insect_data
```